---
title: "RcppR6 examples"
author: "Rich FitzJohn"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RcppR6 examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette builds on the introduction one to provide a
marginally more useful example, and demonstrate a few more features
of RcppR6.

The code used here is a demo package called `examples`, available
within RcppR6 (`system.file("examples/examples",
package="RcppR6")`), and like the `introduction` package doesn't
really do anything that you'd really want to do, or need RcppR6 to
do.

``` {r echo=FALSE, results="asis"}
set.seed(1)
source(system.file("vignette_common.R", package="RcppR6"))
path <- vignette_prepare("examples")
plain_output(tree(path, "introduction"))
```

This package defines three classes, of varying complexity.  It uses
a different way of structuring sources to `introduction`;

``` {r echo=FALSE, results="asis"}
yaml_output(readLines(file.path(path, "inst/RcppR6.yml")))
```

Rather than a single file `inst/RcppR6_classes.yml`, there is a
file `inst/RcppR6.yml` that lists files to include, relative to the
package root.  This isn't necessary, but might help with
organisation.  Each file can define one or more classes -- here
they define a single class each.

Similarly, the definitions are spread over three files:
`inst/include/examples/uniform.hpp`,
`inst/include/examples/stack.hpp` and
`inst/include/examples/empty.hpp`.

## `uniform`

This is a similar example to the Rcpp modules example of a uniform
distribution object.  It's not identical to the modules version.
``` {r echo=FALSE, results="asis"}
cpp_output(readLines(file.path(path, "inst/include/examples/uniform.hpp")))
```

The `Rcpp::RNGScope` bit is probably not needed, actually as Rcpp
attributes will sort that out for us.

The example is declared with a namespace `examples`; unlike
attributes, RcppR6 does not have a problem exporting things that
are not in the global namespace.  This comes in useful for wrapping
library code.

In addition to the class definition, there are also some free
functions defined; `uniform_get_max` and `uniform_set_max`; these
are going to be used to set up active members in the R6 class for
getting and setting the `max` field of the class.  This pattern
(free functions) is useful if you want to do additional error
checking on inputs when calling from R than when calling from C++
(e.g., passing an `int` into a function that expects an unsigned
integer).

Note that the getter takes first argument of type `const Uniform&`;
a const reference to a `Uniform` object.  A non-const reference
would be fine here, as would a *copy* of the object.  The setter
takes first argument `Uniform&`; this needs to be a reference (and
not a pointer).  Passing in a copy here will *appear* to work (as
in; will compile and run) but will not change the value.

The `yaml` for this:
``` {r echo=FALSE, results="asis"}
yaml_output(readLines(file.path(path, "inst/uniform.yml")))
```

In order:

```yaml
uniform:
  name_cpp: examples::Uniform
```

This means that the name of the class on the R side will be
*different* to the name on the C++ side.  We'll export the class as
`uniform` (so that `uniform(...)` will be the constructor in R) but
that the actuall class we are wrapping is called
`examples::Uniform`.  This is how RcppR6 deals with namespaces --
just provide the fully qualified name any time you refer to a
class, function or type.

The next line:

```yaml
  forward_declare: true
```

will arrange to declare (but not define) the class for you.  This
means that we can take less care in writing the package include
file (`inst/include/examples.h`).  In particular, all RcppR6 code
(both `examples/RcppR6_pre.hpp` and `examples/RcppR6_post.hpp`,
along with `Rcpp.h`) can be included before you include
`inst/examples/uniform.hpp` because the class will have been
forward declared.  That's in contrast with the introduction where
we defined the entire class before including
`inst/include/RcppR6_pre.h` so that the `as`/`wrap` templates would
work correctly (see "Extending Rcpp").


Next, up, the constructor:
``` {r echo=FALSE, results="asis"}
yaml <- yaml <- readLines(file.path(path, "inst/uniform.yml"))
i_constructor <- grep("\\s+constructor:", yaml)[[1]]
i_methods <- grep("\\s+methods:", yaml)[[1]]
i_active <- grep("\\s+active:", yaml)[[1]]
yaml_output(yaml[i_constructor:(i_methods - 1)])
```

The first argument here, `roxygen` defines some roxygen content to
include in the generated `R/RcppR6.R` file, but without the leading
`#'`.  This will generate a very minimal set of documentation with
the title, parameters (`min` and `max`) and arrange to `@export`
the object so it appears in the package `NAMESPACE`.  Use of this
field is optional, and will generally require yaml's pipe syntax to
indicate whitespace should be retained in the multiline string.

The `args` field is a yaml ordered map of two arguments.  Both are
`double`s, and both have default values that will be added to the
generated R code:

```r
uniform <- function(min=0.0, max=1.0) { ... }
```

There are two methods `draw` and `range`:
``` {r echo=FALSE, results="asis"}
yaml_output(yaml[i_methods:(i_active - 1)])
```

The `draw` method takes a single integer and returns a
`Rcpp::NumericVector`.  Because no `name_cpp` is given, RcppR6 will
assume that there is a method `draw` within the class that can be
used.  And because no `access` is given RcppR6 assumes that `draw`
is a method and not a free function.

The `range` method calls the free function
`examples::uniform_range()`.  The C++ function takes the argument
`const Uniform& w` but this argument is *not* referred to in the
yaml (the first argument of a free function must take a reference
to the object).  We have to tell RcppR6 that the function is free
(rather than a member) with `access: function` and the name of the
function `name_cpp: examples::uniform_range`.


There are a bunch of active methods, because they're a bit more
varied in the options that they can take
``` {r echo=FALSE, results="asis"}
yaml_output(yaml[i_active:length(yaml)])
```

First, `min` and `max` are direct field accessors.  I've made them
read-only by adding `readonly: true`.  Without this (by default)
they would be read-write.  You can also use `name_cpp` here to
access a different named field within the C++ class than the name
of the R field that will be generated.

The field `the_min` *also* accesses the min field, but does so
through the member function `get_min`.  The `name_cpp_set` field
indicates the name of the setter (`set_min`).  Without providing
this, the field would be read-only.

The field `the_max` does the same thing as `the_min`, but for the
`max` field and uses a pair of free functions
(`examples::uniform_get_max` and `examples::uniform_set_max`) to
achive this.

The active field `u` will return a single random number by calling
the function `examples::draw1()`.

Running RcppR6 (this will create other two classes not yet
discussed)
``` {r }
RcppR6::install(path)
```

Run `devtools::document` to create the `NAMESPACE` file
``` {r }
devtools::document(path)
```

And load the package:
``` {r }
devtools::load_all(path)
```

We can create a `uniform` object:
``` {r }
u <- uniform()
u
```

Draw 10 random numbers:
``` {r }
u$draw(10)
```

Or just one:
``` {r }
u$u
```

The minimum was set to zero and max as one by default:
``` {r }
u$min
u$max
args(uniform)
```

These are read-only:
``` {r error=TRUE}
u$min <- 100
u$max <- 200
```

These can be set through the `the_min` and `the_max` fields (which
are totally redundant here and included only for demonstration)
``` {r }
u$the_min <- 10
u$the_max <- 20
```

new values set:
``` {r }
u$the_min
u$the_max
```

Random number in new range:
``` {r }
u$u
```

## `stack`

This example shows how to wrap a class that is defined elsewhere --
`std::stack` in this case.  It provides an alternative
implementation to the version in the R6 vignette.
``` {r echo=FALSE, results="asis"}
cpp_output(readLines(file.path(path, "inst/include/examples/stack.hpp")))
```

The comments in the C++ code explain largely what is going on;
there are safe wrappers around `pop` and `top` that prevent
crashes.  Better behavour for `top` on an empty stack might be to
throw an error (though that will cause problems as an active member
with R6 < 2.0.0.9000, which includes the current version on CRAN -
2.0.0 at the time of writing).

Then yaml that goes along with this:
``` {r echo=FALSE, results="asis"}
yaml_output(readLines(file.path(path, "inst/stack.yml")))
```

There's not that much more here than for `uniform`:

* `differs` shows how to wrap an operator (though turning this into
   something that dispatches nicely on the R side will take [more
   work](https://github.com/wch/s3ops)
* the class was defined elsewhere as a templated library function
  but we can still wrap it easily enough.

``` {r }
s <- stack()
```

Empty stack has a missing top:
``` {r }
s$top
```

and throws an error when popped (and does not crash!)
``` {r error=TRUE}
s$pop()
```

Push some numbers on the stack:
``` {r }
s$push(1)
s$push(10)
s$push(100)
```

Three things on the stack:
``` {r }
s$size
```

First one is:
``` {r }
s$top
```

`std::stack` does not return on `pop`, unlike Python's stack
``` {r }
s$pop()

s$top
```

empty out the stack by popping repeatedly:
``` {r }
while (!s$empty) {
  s$pop()
}
s$size
```

## `empty`

`empty` is the *simplest posssible* RcppR6 class, defined within
the `simple.hpp` header file:

``` {r echo=FALSE, results="asis"}
cpp_output(readLines(file.path(path, "inst/include/examples/empty.hpp")))
```

This class defines no methods, no constructors, no fields.  It is
totally useless.  But we can still wrap it up.
``` {r }
yaml_output(readLines(file.path(path, "inst/empty.yml")))
```

This probably serves no benefit at all.
``` {r }
e <- empty()
e
```
