---
title: "Introduction to RcppR6"
author: "Rich FitzJohn"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to RcppR6}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

RcppR6 is a code generation approach for exposing C++ classes as R
classes.  It makes it possible to define a class in C++ and expose
it in R as an [R6](https://github.com/wch/R6) class, with reference
semantics for the class on the R side.

This first example walks through setting up the example used in the
package [README](https://github.com/richfitz/RcppR6).  This example
creates a simple "circle" class to show a few features of the
package.

``` {r echo=FALSE}
path <- system.file("examples/README", package="RcppR6")
path <- RcppR6:::prepare_temporary(path, tempfile())
unlink(file.path(path, "tests"), recursive=TRUE)
descr <- readLines(file.path(path, "DESCRIPTION"))
descr <- descr[!grepl("Suggests: testthat", descr, fixed=TRUE)]
writeLines(descr, file.path(path, "DESCRIPTION"))
lang_output <- function(x, lang) {
  cat(c(sprintf("```%s", lang), x, "```"), sep="\n")
}
cpp_output <- function(x) lang_output(x, "c++")
r_output <- function(x) lang_output(x, "r")
yaml_output <- function(x) lang_output(x, "yaml")
plain_output <- function(x) lang_output(x, "plain")
yaml_load <- RcppR6:::yaml_load
```

``` {r echo=FALSE, results="asis"}
cls <- readLines(file.path(path, "inst/include/README.h"))
i1 <- grep("^#include", cls)[[1]]
i2 <- grep("^};", cls)[[1]]
cpp_output(cls[i1:i2])
```

What we want is some way if exposing this class to R so that we
could interact with the methods directly.  The usual approach with
annotating with
```
// [[Rcpp::export]]
```
won't work here because that only wraps free functions; we need a
way to generate these free functions and to organise getting an
object that will keep the state of the class (here, the `radius`
field).

RcppR6 currently requires working in a package structure: this is
encouraged in the [Rcpp intro](TODO:LINK) anyway.  Eventually it
might support inline use, but that is not currently supported.

RcppR6 also requires a particular structure to a package (this
might change!) because it needs to be able to `#include` all the
class definitions in ways that plain Rcpp attributes don't need.

``` {r echo=FALSE, results="asis"}
tree <- function(path, header=path) {
  paste1 <- function(a, b) {
    paste(rep_len(a, length(b)), b)
  }
  indent <- function(x, files) {
    paste0(if (files) "| " else "  ", x)
  }
  is_directory <- function(x) {
    unname(file.info(x)[, "isdir"])
  }
  prefix_file <- "|--="
  prefix_dir  <- "|-+="

  files <- dir(path)
  files_full <- file.path(path, files)
  isdir <- is_directory(files_full)

  ret <- as.list(c(paste1(prefix_dir, files[isdir]),
                   paste1(prefix_file, files[!isdir])))
  files_full <- c(files_full[isdir], files_full[!isdir])
  isdir <- c(isdir[isdir], isdir[!isdir])

  n <- length(ret)
  ret[[n]] <- sub("|", "\\", ret[[n]], fixed=TRUE)
  tmp <- lapply(which(isdir), function(i)
    c(ret[[i]], indent(tree(files_full[[i]], NULL), !all(isdir))))
  ret[isdir] <- tmp

  c(header, unlist(ret))
}
plain_output(tree(path, "README"))
```

(note that the *package* here is called README, as this is the
package used in generating the README for RcppR6 - OK that's
probably confusing).

A file `<package_name>.h` is required within `inst/include`; this
file must include *all* class definitions that RcppR6 is to wrap.
So for this package `README`, the above definition is included
there.

The main work is in the file `inst/RcppR6_classes.yml` this is a
[yaml](http:/yaml.org) file containing key/value pairs indicating
how to export the class.
``` {r echo=FALSE, results="asis"}
yaml <- readLines(file.path(path, "inst/RcppR6_classes.yml"))
yaml_output(yaml)
```

This is the main file that needs editing to organise export of the
class.  Compared with Rcpp modules it contains a lot of type
information that could have been lifted from the class definition,
but future versions may remove this limitation.

The yaml is hopefully fairly self-explanatory, and illustrates
*almost* all of the features needed features.  Below, I'll walk
through each section in turn: the constructor, the method, and then
the "active" fields.

First, a **constructor**; this is the function that creates the
object.

``` {r echo=FALSE, results="asis"}
i_constructor <- grep("\\s+constructor:", yaml)[[1]]
i_methods <- grep("\\s+methods:", yaml)[[1]]
i_active <- grep("\\s+active:", yaml)[[1]]
yaml_output(yaml[i_constructor:(i_methods - 1)])
```

The `args` element is a yaml ordered map (by definition a map in
yaml is not ordered, but ordering is critical here).  Each pair in
the args is of the form: `<name>: <type>`, so this just says that
there is one argument, `radius`, which has type `double`.  Note
that the name need not match up with the name in the C++ class (and
does not here) but the name must be valid in *both* R and C++.
That means no dots are allowed.

If the type contains colons (e.g., `std::string`) it will probably
be required to enclose the type in double quotes or the yaml parser
will throw an error.

An alternative way of writing the arguments using an ordered map in
yaml is:

```yaml
    args:
      - {radius: double}
```

with additional arguments as additional list elements (`-`).

By default (and above) this will use the constructor of the C++
class, but it is possible to specify a free function that returns
an object of class `circle`.  So if you had:

```c++
circle make_circle(double radius) {
  return circle(radius);
}
```

you could use that by writing:

```yaml
  constructor:
    name_cpp: make_circle
    args: ... # as above
```

In the generated object, a function with the name of the class will
be generated (so `circle`) taking the one argument `radius`.  (Note
that this differs from the usual way that R6 objects are
generated).

```r
circle(1.0) # would create a circle with radius 1.0
```

Next, one **method** is defined in the yaml:
``` {r echo=FALSE, results="asis"}
yaml_output(yaml[i_methods:(i_active - 1)])
```

The `methods` field contains any number of methods; here, only the
method `area` is defined.

Each method may contain an `args` element (as the constructor did)
but here it is omitted because the function has no arguments.

The field `return_type` is required; here the method is going to
return a `double`.

If `x` is a circle object, created by the constructor `circle`
above, then the method would be used by writing:

```r
r$area()
```

Instead of using class methods, RcppR6 can also use functions that
take references to objects and turn these into methods in the
generated object.  So if we had a function:

```c++
double circle_area(const circle& x) {
  return x.area();
}
```

that could be specified above by writing:

```yaml
  area:
    return_type: double
    access: function
    name_cpp: circle_area
```

The field `access` takes values "member" (the default, indicating a
member function) or `function`, indicating a free function.  The
field `name_cpp` is the name of the function (or member) in C++.
The actual name `area` will be used in the generated object.

Finally, two **active fields** are specified; `circumference` and
`radius`:
``` {r echo=FALSE, results="asis"}
yaml_output(yaml[i_active:length(yaml)])
```

The first, `circumference` provides both setters and getters
(`name_cpp` and `name_cpp_set`, respectively).  The getter could be
ommited here because it shares the same name as the R name for the
field.

Getters must be a function that takes no arguments and return a
thing that they get, while setters must take a single thing and
return void.  It's probably a good idea for a getter to be a const
method (see the C++ definition of `circumference` above), but
that's not enforced.  A setter can do whatever argument checking it
wants.

"active" members also require a `type` entry: this is the return
type for the getter and the argument type for the setter.

The `circumference` field has access "member"; this means it is
accessed by member functions).  The other alternatives are
"function" (as for the "methods" section above) and `field`.

The `radius` active member has access `field`: it means that the
field is accesed directly, with no argument checking.  By default
this is a read-write binding, but by specifying `readonly: true`,
this can be made read-only.  Note here that a shorthand `{...}`
notation is used - this is equivalent yaml to:
``` {r echo=FALSE, results="asis"}
dat <- RcppR6:::yaml_load(paste(yaml, collapse="\n"))
yaml_output(yaml::as.yaml(dat$circle$active["radius"]))
```

That's almost all the bits: there are a few other required bits for
the package header:
``` {r echo=FALSE, results="asis"}
tmp <- grep("^//", cls, value=TRUE, invert=TRUE)
cpp_output(gsub("\n\n+", "\n\n", paste(tmp, collapse="\n")))
```

In addition to the class defnition above, there are a few extra
bits:

* header guards (optional, but probably going to be needed)
* including the file `<README/RcppR6_pre.hpp>`
    - this will be added to the `inst/include/README` directory when
      running RcppR6; it contains prototypes for the `as` and `wrap`
      functions required to export types from C++ to R (see the
      "extending Rcpp" vignette).
    - this needs to be included after your classes have been
      *declared*, but may be included before your classes have been
      *defined*.  It must be included *after* `<RcppCommon.h>` and
      *before* `<Rcpp.h>`.
* including the file <README/RcppR6_post.hpp>`
    - this will include the definition of the `as` and `wrap`
      functions, as well as `<Rcpp.h>` (if it hasn't already been
      included) and some support code needed by RcppR6.

The `DESCRIPTION` file contains nothing special:
``` {r echo=FALSE, results="asis"}
plain_output(readLines(file.path(path, "DESCRIPTION")))
```

and the NAMESPACE file is empty.

With everything in place, let's go:
``` {r }
RcppR6::install(path)
```

Quite a few files have been added, and some of the existing files
have been updated
``` {r echo=FALSE, results="asis"}
plain_output(tree(path, "README"))
```

RcppR6 reads the DESCRIPTION and adds the required packages: the
package must import Rcpp and R6, and must include Rcpp in
`LinkingTo:`.  Note that RcppR6 *does not appear anywhere* in the
description: once RcppR6 has generated code, it is done and is not
a dependency.
``` {r echo=FALSE, results="asis"}
plain_output(readLines(file.path(path, "DESCRIPTION")))
```

Because `src/Makevars` was missing originally, it has been creted
with contents:
``` {r echo=FALSE, results="asis"}
plain_output(readLines(file.path(path, "src/Makevars")))
```

(if it already existed it would have been left alone and you would
have to add this yourself).

The other files: `inst/include/README/RcppR6_pre.hpp`,
`inst/include/README/RcppR6_post.hpp`,
`inst/include/README/RcppR6_support.hpp`, `R/RcppR6.R` and
`src/RcppR6.R` contain boilerplate glue code, and the files
`R/RcppExports.R` and `src/RcppExports.cpp` contain the usual Rcpp
attributes generated code to support them.  See the bottom of this
file for what is generated, if you are curious.

The package can now be compiled.  I'm using `devtools::document`
here because the generated code includes enough roxygen hints to
generate a minimal `NAMESPACE` file:
``` {r }
devtools::document(path)
```

The package can now be loaded:
``` {r }
devtools::load_all(path)
```

and a circle object can be created:
``` {r }
obj <- circle(1.0)
obj
```

This object has class `r class(obj)[[1]]`:
``` {r }
class(obj)
```

There's a data member `.ptr`
``` {r }
obj$.ptr
```

This is the actual external pointer object that holds a reference
to the underlying C++ class instance.  It is not designed to be
interacted with directly.  The `initialize` method is also not
meant to be used directly:
``` {r }
obj$initialize
```

(this is used by the support code and why the objects are not
created with the usual R6 `circle$new()` syntax).

The one method and two active binding members are present in the
class: `area`, `circumference` and `radius`.

The area of a circle of radius 1 is of course pi
``` {r }
obj$area()
obj$area() - pi
```

Because the radius is an active member, parentheses are not needed
to access it:
``` {r }
obj$radius
```

and it can be set as if it were a field:
``` {r }
obj$radius <- 2
obj$radius
obj$area()
```

Similarly, the circumference can be set, routing through the
`set_circumference` function that converts to a radius:
``` {r }
obj$circumference <- 1.0
obj$circumference
obj$radius
```

and because that function has error checking in it, you can't set
negative values:
``` {r error=TRUE}
obj$circumference <- -1
```

Once set up, RcppR6 should be relatively cheap to run as it detects
that nothing has changed:
``` {r }
system.time(RcppR6::RcppR6(path))
```

# Contents of generated files:

`inst/include/README/RcppR6_pre.hpp`:
``` {r echo=FALSE, results="asis"}
cpp_output(readLines(file.path(path, "inst/include/README/RcppR6_pre.hpp")))
```

`inst/include/README/RcppR6_post.hpp`:
``` {r echo=FALSE, results="asis"}
cpp_output(readLines(file.path(path, "inst/include/README/RcppR6_post.hpp")))
```

`inst/include/README/RcppR6_support.hpp`:
``` {r echo=FALSE, results="asis"}
cpp_output(readLines(file.path(path, "inst/include/README/RcppR6_support.hpp")))
```

`R/RcppR6.R`:
``` {r echo=FALSE, results="asis"}
r_output(readLines(file.path(path, "R/RcppR6.R")))
```

`R/RcppR6.R`:
``` {r echo=FALSE, results="asis"}
cpp_output(readLines(file.path(path, "src/RcppR6.cpp")))
```
